---
title: "Building Tidy Data"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo=TRUE)
```

# Overview

This notebook serves as reference for the tidying procedures used to prepare the
study data for analysis.

# Forming the Data Frame

## Loading the Raw Data

Load and verify the raw owner data.

```{r}
df_owners <- read.csv('../data/raw-owners.csv', header=TRUE, skipNul=TRUE)
names(df_owners)[names(df_owners) == 'What.is.your.email.'] <- 'owner_id'
stopifnot(identical(dim(df_owners)+0, c(717, 6)))
str(df_owners)
```

Load and verify the raw dog data.

```{r}
df_dogs <- read.csv('../data/raw-dogs.csv', header=TRUE, skipNul=TRUE)
names(df_dogs)[names(df_dogs) == 'id'] <- 'owner_id'
stopifnot(identical(dim(df_dogs)+0, c(969, 45)))
str(df_dogs[, c(1:3, 43:45)])
```

## Forming Complete Responses

```{r}
# Merge the raw owners and dogs data on the owner id.
df <- merge(df_owners, df_dogs, by='owner_id', all=TRUE)
stopifnot(identical(dim(df)+0, c(1181, 50)))
str(df[, c(1:3, 47:50)])
```

## Applying Readable Column Names

```{r}
colnames(df) <- readr::read_lines('../docs/readable-column-names.csv')
str(df)
```

# Dropping Unnecessary Columns

We use the following section to drop unnecessary columns from the data frame.

```{r}
df <- df %>%
  select(-(owner_referral_source:owner_token)) %>%
  select(-(owner_has_another_dog_to_enter:dog_token)) 
stopifnot(identical(dim(df)+0, c(1181, 41)))
str(df)
```

# Adjusting Data Types

We use the following sections to adjust field types in the data frame.

```{r}

#copy data frame
df_adj_dtypes <- df

# change comma-separated lists into separate columns
# separate each answer into its own column.
# convert to factor TRUE / FALSE

#dog_adolescent_or_adult_training_trainer
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(trainer_group_classes = as.factor(ifelse(
    dog_adolescent_or_adult_training_trainer == "" | 
      dog_adolescent_or_adult_training_trainer == "I don't know", NA, ifelse(
      grepl('Myself/family member attended group classes', 
            dog_adolescent_or_adult_training_trainer), TRUE, FALSE)))) %>%
  mutate(trainer_boot_camp = as.factor(ifelse(
    dog_adolescent_or_adult_training_trainer == "" | 
      dog_adolescent_or_adult_training_trainer == "I don't know", NA, ifelse(
      grepl('Sent to professional training school/boot camp', 
            dog_adolescent_or_adult_training_trainer), TRUE, FALSE)))) %>%
  mutate(trainer_home_private = as.factor(ifelse(
    dog_adolescent_or_adult_training_trainer == "" | 
      dog_adolescent_or_adult_training_trainer == "I don't know", NA, ifelse(
      grepl('Hired in home private trainer', dog_adolescent_or_adult_training_trainer), TRUE, FALSE))))


#dog_adolescent_or_adult_training_completion_age
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(training_complete_age_6_to_18mo = as.factor(ifelse(
    dog_adolescent_or_adult_training_completion_age == "" | 
      dog_adolescent_or_adult_training_completion_age == "I don't know", NA, ifelse(
      grepl('6 to 18 months \\(adolescence)', 
            dog_adolescent_or_adult_training_completion_age), TRUE, FALSE)))) %>%
  mutate(training_complete_age_18_to_36mo = as.factor(ifelse(
    dog_adolescent_or_adult_training_completion_age == "" | 
      dog_adolescent_or_adult_training_completion_age == "I don't know", NA, ifelse(
      grepl('18-36 months \\(social maturity)', 
            dog_adolescent_or_adult_training_completion_age), TRUE, FALSE)))) %>%
  mutate(training_complete_age_continue_in_home = as.factor(ifelse(
    dog_adolescent_or_adult_training_completion_age == "" | 
      dog_adolescent_or_adult_training_completion_age == "I don't know", NA, ifelse(
      grepl('Continue to practice and use training skills in home',
            dog_adolescent_or_adult_training_completion_age), TRUE, FALSE)))) %>%
  mutate(training_complete_age_continue_classes = as.factor(ifelse(
    dog_adolescent_or_adult_training_completion_age == "" | 
      dog_adolescent_or_adult_training_completion_age == "I don't know", NA, ifelse(
      grepl('Continue to attend classes periodically \\(nose work, agility, trick training, etc\\.)',
            dog_adolescent_or_adult_training_completion_age), TRUE, FALSE))))
  

#dog_agression_scenario
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(aggression_strange_people_in_home = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Strangers visiting the home', 
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_strange_people_away_from_home = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Strangers away from the home', 
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_familiar_people_in_home = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Familiar people in the home',
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_familiar_people_away_from_home = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Familiar people away from the home',
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_strange_dog_in_home = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Unfamiliar dogs in the home', 
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_strange_dog_away_from_home = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Unfamiliar dogs away from the home', 
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_familiar_dog_in_home = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Familiar dogs in the home',
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_familiar_dog_away_from_home = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Familiar dogs away from the home',
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_vet = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Veterinarians',
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_other_animals = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Animals other than dogs',
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_groomer = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Groomers',
            dog_aggression_scenario), TRUE, FALSE)))) %>%
  mutate(aggression_trainer = as.factor(ifelse(
    dog_aggression_scenario == "" | 
      dog_aggression_scenario == "I don't know", NA, ifelse(
      grepl('Trainers',
            dog_aggression_scenario), TRUE, FALSE))))

# need 'other' category


#dog_fear_anxiety_scenario
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(fear_anxiety_crowds = as.factor(ifelse(
    dog_fear_anxiety_scenario == "" | 
      dog_fear_anxiety_scenario == "I don't know", NA, ifelse(
      grepl('Fear of crowds', 
            dog_fear_anxiety_scenario), TRUE, FALSE)))) %>%
  mutate(fear_anxiety_noises = as.factor(ifelse(
    dog_fear_anxiety_scenario == "" | 
      dog_fear_anxiety_scenario == "I don't know", NA, ifelse(
      grepl('Fear of noises', 
            dog_fear_anxiety_scenario), TRUE, FALSE)))) %>%
  mutate(fear_anxiety_thunderstorms = as.factor(ifelse(
    dog_fear_anxiety_scenario == "" | 
      dog_fear_anxiety_scenario == "I don't know", NA, ifelse(
      grepl('Fear of thunderstorms',
            dog_fear_anxiety_scenario), TRUE, FALSE)))) %>%
  mutate(fear_anxiety_vet = as.factor(ifelse(
    dog_fear_anxiety_scenario == "" | 
      dog_fear_anxiety_scenario == "I don't know", NA, ifelse(
      grepl('Fear of veterinary visits', 
            dog_fear_anxiety_scenario), TRUE, FALSE))))  %>%
  mutate(fear_anxiety_other_dogs = as.factor(ifelse(
    dog_fear_anxiety_scenario == "" | 
      dog_fear_anxiety_scenario == "I don't know", NA, ifelse(
      grepl('Fear of other dogs',
            dog_fear_anxiety_scenario), TRUE, FALSE))))  %>%
  mutate(fear_anxiety_other_animals = as.factor(ifelse(
    dog_fear_anxiety_scenario == "" | 
      dog_fear_anxiety_scenario == "I don't know", NA, ifelse(
      grepl('Fear of animals other than dogs', 
            dog_fear_anxiety_scenario), TRUE, FALSE)))) %>%
  mutate(fear_anxiety_separation = as.factor(ifelse(
    dog_fear_anxiety_scenario == "" | 
      dog_fear_anxiety_scenario == "I don't know", NA, ifelse(
      grepl('Separation anxiety',
            dog_fear_anxiety_scenario), TRUE, FALSE)))) %>%
  mutate(fear_anxiety_travel = as.factor(ifelse(
    dog_fear_anxiety_scenario == "" | 
      dog_fear_anxiety_scenario == "I don't know", NA, ifelse(
      grepl('Travel anxiety',
            dog_fear_anxiety_scenario), TRUE, FALSE)))) %>%
  mutate(fear_anxiety_general = as.factor(ifelse(
    dog_fear_anxiety_scenario == "" | 
      dog_fear_anxiety_scenario == "I don't know", NA, ifelse(
      grepl('Generalized anxiety',
            dog_fear_anxiety_scenario), TRUE, FALSE)))) %>%
  mutate(fear_anxiety_ptsd = as.factor(ifelse(
    dog_fear_anxiety_scenario == "" | 
      dog_fear_anxiety_scenario == "I don't know", NA, ifelse(
      grepl('Post-traumatic Stress Disorder \\(PTSD)',
            dog_fear_anxiety_scenario), TRUE, FALSE)))) 


#problematic jumping subject
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(jumping_subject_familiar_people = as.factor(ifelse(
    dog_problematic_jump_subject == "" | 
      dog_problematic_jump_subject == "I don't know", NA, ifelse(
      grepl('Familiar people', 
            dog_problematic_jump_subject), TRUE, FALSE)))) %>%
  mutate(jumping_subject_owners = as.factor(ifelse(
    dog_problematic_jump_subject == "" | 
      dog_problematic_jump_subject == "I don't know", NA, ifelse(
      grepl('Owners', 
            dog_problematic_jump_subject), TRUE, FALSE)))) %>%
  mutate(jumping_subject_strangers = as.factor(ifelse(
    dog_problematic_jump_subject == "" | 
      dog_problematic_jump_subject == "I don't know", NA, ifelse(
      grepl('Stangers',
            dog_problematic_jump_subject), TRUE, FALSE)))) 
  
  
#excessive barking scenario
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(excessive_barking_inside_triggers = as.factor(ifelse(
    dog_excessive_barking_scenario == "" | 
      dog_excessive_barking_scenario == "I don't know", NA, ifelse(
      grepl('At triggers \\(inside)', 
            dog_excessive_barking_scenario), TRUE, FALSE)))) %>%
  mutate(excessive_barking_outside_triggers = as.factor(ifelse(
    dog_excessive_barking_scenario == "" | 
      dog_excessive_barking_scenario == "I don't know", NA, ifelse(
      grepl('At triggers \\(outside)', 
            dog_excessive_barking_scenario), TRUE, FALSE)))) %>%
  mutate(excessive_barking_owner_present = as.factor(ifelse(
    dog_excessive_barking_scenario == "" | 
      dog_excessive_barking_scenario == "I don't know", NA, ifelse(
      grepl('Owners are present',
            dog_excessive_barking_scenario), TRUE, FALSE)))) %>%
  mutate(excessive_barking_owner_away = as.factor(ifelse(
    dog_excessive_barking_scenario == "" | 
      dog_excessive_barking_scenario == "I don't know", NA, ifelse(
      grepl('Owners are away', 
            dog_excessive_barking_scenario), TRUE, FALSE))))  %>%
  mutate(excessive_barking_attention = as.factor(ifelse(
    dog_excessive_barking_scenario == "" | 
      dog_excessive_barking_scenario == "I don't know", NA, ifelse(
      grepl('To get attention',
            dog_excessive_barking_scenario), TRUE, FALSE))))  %>%
  mutate(excessive_barking_car_rides = as.factor(ifelse(
    dog_excessive_barking_scenario == "" | 
      dog_excessive_barking_scenario == "I don't know", NA, ifelse(
      grepl('During car rides', 
            dog_excessive_barking_scenario), TRUE, FALSE))))
  

#coprophagia type
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(coprophagia_own_feces = as.factor(ifelse(
    dog_coprophagia_type == "" | 
      dog_coprophagia_type == "I don't know", NA, ifelse(
      grepl('Their own feces', 
            dog_coprophagia_type), TRUE, FALSE)))) %>%
  mutate(coprophagia_other_dogs = as.factor(ifelse(
    dog_coprophagia_type == "" | 
      dog_coprophagia_type == "I don't know", NA, ifelse(
      grepl('Dog feces other than their own', 
            dog_coprophagia_type), TRUE, FALSE)))) %>%
  mutate(coprophagia_other_animals = as.factor(ifelse(
    dog_coprophagia_type == "" | 
      dog_coprophagia_type == "I don't know", NA, ifelse(
      grepl('The feces of other animals',
            dog_coprophagia_type), TRUE, FALSE)))) 


#dog_repetitive_behavior_type
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(repetitive_behavior_collecting_objects = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Collecting/arranging objects', 
            dog_repetitive_behavior_type), TRUE, FALSE)))) %>%
  mutate(repetitive_behavior_licking_wrist = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Licking of the wrist/hock', 
            dog_repetitive_behavior_type), TRUE, FALSE)))) %>%
  mutate(repetitive_behavior_sucking_flank = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Sucking flank regions/blankets',
            dog_repetitive_behavior_type), TRUE, FALSE)))) %>%
  mutate(repetitive_behavior_ball_fetish = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Tennis ball fetish', 
            dog_repetitive_behavior_type), TRUE, FALSE))))  %>%
  mutate(repetitive_behavior_running_in_patterns = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Running in geometric patterns',
            dog_repetitive_behavior_type), TRUE, FALSE))))  %>%
  mutate(repetitive_behavior_digging = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Digging in the yard', 
            dog_repetitive_behavior_type), TRUE, FALSE)))) %>%
  mutate(repetitive_behavior_nail_biting = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Nail biting',
            dog_repetitive_behavior_type), TRUE, FALSE)))) %>%
  mutate(repetitive_behavior_fly_snapping = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Fly snapping \\(when no flies are around)',
            dog_repetitive_behavior_type), TRUE, FALSE)))) %>%
  mutate(repetitive_behavior_shadow_chasing = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Shadow/light chasing',
            dog_repetitive_behavior_type), TRUE, FALSE)))) %>%
  mutate(repetitive_behavior_stone_chewing = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Stone/rock chewing',
            dog_repetitive_behavior_type), TRUE, FALSE)))) %>%
  mutate(repetitive_behavior_tail_chasing = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Tail chasing',
            dog_repetitive_behavior_type), TRUE, FALSE)))) %>%
  mutate(repetitive_behavior_spinning = as.factor(ifelse(
    dog_repetitive_behavior_type == "" | 
      dog_repetitive_behavior_type == "I don't know", NA, ifelse(
      grepl('Spinning',
            dog_repetitive_behavior_type), TRUE, FALSE)))) 


#house soiling scenario
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(house_soiling_excited = as.factor(ifelse(
    dog_house_soil_scenario == "" | 
      dog_house_soil_scenario == "I don't know", NA, ifelse(
      grepl('Exited/overwhelmed', 
            dog_house_soil_scenario), TRUE, FALSE)))) %>%
  mutate(house_soiling_owner_present = as.factor(ifelse(
    dog_house_soil_scenario == "" | 
      dog_house_soil_scenario == "I don't know", NA, ifelse(
      grepl('Owners are present', 
            dog_house_soil_scenario), TRUE, FALSE)))) %>%
  mutate(house_soiling_owner_away = as.factor(ifelse(
    dog_house_soil_scenario == "" | 
      dog_house_soil_scenario == "I don't know", NA, ifelse(
      grepl('Owners are away',
            dog_house_soil_scenario), TRUE, FALSE)))) 


#house soiling type
# still need to reconcile both urine and feces answers
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(house_soiling_feces = as.factor(ifelse(
    dog_house_soil_type == "" | 
      dog_house_soil_type == "I don't know", NA, ifelse(
      grepl('Feces', 
            dog_house_soil_type), TRUE, FALSE)))) %>%
  mutate(house_soiling_urine = as.factor(ifelse(
    dog_house_soil_type == "" | 
      dog_house_soil_type == "I don't know", NA, ifelse(
      grepl('Urine', 
            dog_house_soil_type), TRUE, FALSE)))) %>%
  mutate(house_soiling_both = as.factor(ifelse(
    dog_house_soil_type == "" | 
      dog_house_soil_type == "I don't know", NA, ifelse(
      grepl('Both feces and urine',
            dog_house_soil_type), TRUE, FALSE)))) 


#house soiling location
# need to add "other" category
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(house_soiling_anywhere = as.factor(ifelse(
    dog_house_soil_location == "" | 
      dog_house_soil_location == "I don't know", NA, ifelse(
      grepl('Anywhere', 
            dog_house_soil_location), TRUE, FALSE)))) %>%
  mutate(house_soiling_specific_locations = as.factor(ifelse(
    dog_house_soil_location == "" | 
      dog_house_soil_location == "I don't know", NA, ifelse(
      grepl('Specific locations', 
            dog_house_soil_location), TRUE, FALSE)))) 


#rolling in repulsive material type
# add other type?
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(repulsive_material_urine = as.factor(ifelse(
    dog_repulsive_material_type == "" | 
      dog_repulsive_material_type == "I don't know", NA, ifelse(
      grepl('Urine', 
            dog_repulsive_material_type), TRUE, FALSE)))) %>%
  mutate(repulsive_material_feces = as.factor(ifelse(
    dog_repulsive_material_type == "" | 
      dog_repulsive_material_type == "I don't know", NA, ifelse(
      grepl('Feces', 
            dog_repulsive_material_type), TRUE, FALSE)))) %>%
  mutate(repulsive_material_dead_stuff = as.factor(ifelse(
    dog_repulsive_material_type == "" | 
      dog_repulsive_material_type == "I don't know", NA, ifelse(
      grepl('Dead stuff',
            dog_repulsive_material_type), TRUE, FALSE)))) %>%
  mutate(repulsive_material_garbage = as.factor(ifelse(
    dog_repulsive_material_type == "" | 
      dog_repulsive_material_type == "I don't know", NA, ifelse(
      grepl('Garbage',
            dog_repulsive_material_type), TRUE, FALSE))))


#hyperactive scenario
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(hyperactive_scenario_constant_moving = as.factor(ifelse(
    dog_hyperactive_scenario == "" | 
      dog_hyperactive_scenario == "I don't know", NA, ifelse(
      grepl('Constant moving/jumping', 
            dog_hyperactive_scenario), TRUE, FALSE)))) %>%
  mutate(hyperactive_scenario_difficulty_settling = as.factor(ifelse(
    dog_hyperactive_scenario == "" | 
      dog_hyperactive_scenario == "I don't know", NA, ifelse(
      grepl('Difficulty settling', 
            dog_hyperactive_scenario), TRUE, FALSE)))) %>%
  mutate(hyperactive_scenario_easily_distracted = as.factor(ifelse(
    dog_hyperactive_scenario == "" | 
      dog_hyperactive_scenario == "I don't know", NA, ifelse(
      grepl('Easily distracted',
            dog_hyperactive_scenario), TRUE, FALSE)))) %>%
  mutate(hyperactive_scenario_impulsive = as.factor(ifelse(
    dog_hyperactive_scenario == "" | 
      dog_hyperactive_scenario == "I don't know", NA, ifelse(
      grepl('Impulsiveness',
            dog_hyperactive_scenario), TRUE, FALSE))))


#destruction scenario
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(destruction_scenario_owner_present = as.factor(ifelse(
    dog_destruction_scenario == "" | 
      dog_destruction_scenario == "I don't know", NA, ifelse(
      grepl('Owner is present', 
            dog_destruction_scenario), TRUE, FALSE)))) %>%
  mutate(destruction_scenario_owner_away = as.factor(ifelse(
    dog_destruction_scenario == "" | 
      dog_destruction_scenario == "I don't know", NA, ifelse(
      grepl('Owner is away', 
            dog_destruction_scenario), TRUE, FALSE)))) 


#escape scenario
# other category?
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(escape_scenario_when_out = as.factor(ifelse(
    dog_escape_scenario == "" | 
      dog_escape_scenario == "I don't know", NA, ifelse(
      grepl('Escaped when out', 
            dog_escape_scenario), TRUE, FALSE)))) %>%
  mutate(escape_scenario_from_home = as.factor(ifelse(
    dog_escape_scenario == "" | 
      dog_escape_scenario == "I don't know", NA, ifelse(
      grepl('Escapes from home', 
            dog_escape_scenario), TRUE, FALSE)))) %>%
  mutate(escape_scenario_from_confinement = as.factor(ifelse(
    dog_escape_scenario == "" | 
      dog_escape_scenario == "I don't know", NA, ifelse(
      grepl('Escapes from confinement',
            dog_escape_scenario), TRUE, FALSE)))) %>%
  mutate(escape_scenario_returns_home = as.factor(ifelse(
    dog_escape_scenario == "" | 
      dog_escape_scenario == "I don't know", NA, ifelse(
      grepl('Returns home after escape',
            dog_escape_scenario), TRUE, FALSE))))

#mount subject
# other category?
df_adj_dtypes <- df_adj_dtypes %>%
  mutate(mount_subject_familiar_dogs = as.factor(ifelse(
    dog_mount_subject == "" | 
      dog_mount_subject == "I don't know", NA, ifelse(
      grepl('Familiar dogs', 
            dog_mount_subject), TRUE, FALSE)))) %>%
  mutate(mount_subject_unfamiliar_dogs = as.factor(ifelse(
    dog_mount_subject == "" | 
      dog_mount_subject == "I don't know", NA, ifelse(
      grepl('Unfamiliar dogs', 
            dog_mount_subject), TRUE, FALSE)))) %>%
  mutate(mount_subject_inanimate_objects = as.factor(ifelse(
    dog_mount_subject == "" | 
      dog_mount_subject == "I don't know", NA, ifelse(
      grepl('Inanimate objects',
            dog_mount_subject), TRUE, FALSE)))) %>%
  mutate(mount_subject_people = as.factor(ifelse(
    dog_mount_subject == "" | 
      dog_mount_subject == "I don't know", NA, ifelse(
      grepl('People',
            dog_mount_subject), TRUE, FALSE))))

```  

```{r}

# replace blank fields with "NA" ( = no answer / missing values)
df_adj_dtypes[df_adj_dtypes==""] <- NA

# convert character-type columns to factors
factors <- c(
  'owner_gender',
  'dog_gender',
  'dog_age_is_known',
  'dog_was_acquired_pre_adolescent',
  'dog_attended_pre_adolescent_training',
  'dog_pre_adolescent_training_type',
  'dog_pre_adolescent_training_location',
  'dog_adolescent_or_adult_training',
  'dog_adolescent_or_adult_training_style',
  'dog_was_aggressive',
  'dog_has_fear_or_anxiety',
  'dog_has_jumped_problematically',
  'dog_has_barked_excessively',
  'dog_has_coprophagia',
  'dog_has_house_soiled',
  'dog_has_rolled_in_repulsive_material',
  'dog_was_hyperactive',
  'dog_was_destructive',
  'dog_has_escaped',
  'dog_has_mounted'
)

for (c in factors) {
  df_adj_dtypes[, c] <- as.factor(df_adj_dtypes[, c])
}

str(df[, factors])

```

### Boolean

Some factor columns can be converted to boolean (i.e., logical).

```{r}
df <- df %>%
  mutate(at_least_1yo = ifelse(at_least_1yo == 'Yes', TRUE, FALSE)) %>%
  mutate(neutered = ifelse(neutered == 'Yes', TRUE, FALSE)) %>%
  mutate(acq_12_wo_or_less = ifelse(
    acq_12_wo_or_less == "I don't know", NA, ifelse(
      acq_12_wo_or_less == 'Yes', TRUE, FALSE)))
```

# Additional Data Cleaning

We use the following sections to carry out additional tidying procedures.

```{r}
# TODO
```

# Final Summary

Take a last look at the data before saving it to disk.

```{r}
dim(df)
summary(df)
```

# Saving the Tidy Data

Save the data to a file in the data directory using RDS format so that the data
types are stored and the resulting file is compressed.

```{r}
saveRDS(df, '../data/tidy.Rds')
```
