---
title: "Inferential Data Analysis"
author: "Liz Keller"
output: html_notebook
date: "`r Sys.Date()`"

---

```{r setup, include=FALSE}
library(tidyverse)
library(ggpubr)
library(GGally)
library(car)

knitr::opts_chunk$set(echo=TRUE)
```

# Loading the Data

Load the raw data and verify its dimensions and structure.

```{r}
df <- readRDS('../data/tidy.Rds')
dim(df)
summary(df)

```

# Control vs Experimental Group

control = no adolescent/adult training, while also controlling for pre-adolescent training

```{r}
df_control <- readRDS('../data/df_control.Rds')
df_exp <- readRDS('../data/df_exp.Rds')
df_no_training <- readRDS('../data/df_no_training.Rds')
df_adult_only <- readRDS('../data/df_adult_only.Rds')

dim(df_control)
dim(df_exp)
dim(df_no_training)
dim(df_adult_only)
#summary(df)

```

#dataset can be spilt into two groups to control for pre-adolescent training status

```{r}

df_preK_training <- rbind(df_control,df_exp)
df_no_preK_training <- rbind(df_adult_only,df_no_training)

nrow(df_preK_training) + nrow(df_no_preK_training)

df_all <- rbind(df_preK_training,df_no_preK_training)

df_valid_responses <- subset(df, dog_adolescent_or_adult_training==TRUE | dog_adolescent_or_adult_training==FALSE)


```

## Overview

The first question we want to answer is if training during adolescence and/or 
adulthood (independent of pre-adolescent / puppy training) has an impact on the 
likelihood of a dog having certain behavior problems. 

The behavior problems we are exploring are:

- Aggression
- Barking excessively
- Coprophagia
- Destructive behavior
- Escaping/running away
- Fear/anxiety
- House soiling
- Hyperactivity
- Jumping problematically
- Mounting
- Repetitive behavior
- Rolling in repulsive materials

```{r}
outcomes <- c(
  'dog_was_aggressive',
  'dog_has_fear_or_anxiety',
  'dog_has_jumped_problematically',
  'dog_has_barked_excessively',
  'dog_has_coprophagia',
  'dog_has_repetitive_behavior',
  'dog_has_house_soiled',
  'dog_has_rolled_in_repulsive_material',
  'dog_was_hyperactive',
  'dog_was_destructive',
  'dog_has_escaped',
  'dog_has_mounted'
)
outcomes <- sort(outcomes)
```

## Statistical Models

We check for correlation between the predictor (attending adolescent / adult training) and
each outcome (the presence of a specific behavior problem). 

Here we perform the Fisher Exact test with a Benjamini-Hochberg correction.
(The Fischer test is useful for categorical data where the sample size is small)

```{r}

pred <- 'dog_adolescent_or_adult_training'

idx <- NULL
p_values <- NULL
odds_ratios <- NULL 
for (outcome in outcomes) {
  tbl <- table(df_valid_responses[, pred], df_valid_responses[, outcome], dnn=c(pred, outcome))
  fisher <- fisher.test(tbl)
  idx <- c(idx, outcome)
  p_values <- c(p_values, fisher$p.value)
  odds_ratios <- c(odds_ratios, fisher$estimate[[1]])
}

# Correct for the possibility of Type I errors.
p_values <- p.adjust(p_values, method='BH')

# Form a result data frame.
df_out <- data.frame(outcome=idx, p_value=p_values,
                     odds_ratio=odds_ratios)

add_sig_columns <- function(df) {
  df$level <- ''
  df[df$p_value <= .05, 'level'] <- '*'
  df[df$p_value <= .01, 'level'] <- '**'
  df[df$p_value <= .001, 'level'] <- '***'
  
  df$dir <- ''
  df[df$odds_ratio < 1, 'dir'] <- '-'
  df[df$odds_ratio > 1, 'dir'] <- '+'
  
  return (df)
}

df_out <- add_sig_columns(df_out)
print(knitr::kable(df_out))

```
There is a weakly significant negative correlation between adult 
training and some specific behaviors: house soiling and destructiveness

there is a weakly significant positive correlation with adult training and
problematic jumping


#check chi-squared test for comparison

```{r}

pred <- 'dog_adolescent_or_adult_training'

idx <- NULL
p_values <- NULL
odds_ratios <- NULL 
for (outcome in outcomes) {
  tbl <- table(df_valid_responses[, pred], df_valid_responses[, outcome], dnn=c(pred, outcome))
  chisq <- chisq.test(tbl)
  idx <- c(idx, outcome)
  p_values <- c(p_values, chisq$p.value)
  #odds_ratios <- c(odds_ratios, chisq$estimate[[1]])
}

# Correct for the possibility of Type I errors (false positive/discovery rate).
p_values <- p.adjust(p_values, method='BH')

# Form a result data frame.
df_out_chisq <- data.frame(outcome=idx, p_value=p_values)

df_out_chisq <- add_sig_columns(df_out_chisq)
print(knitr::kable(df_out_chisq))

```
Result: same variables are of same order of significance with chi squared test


## Binary Logistic Regression

```{r}
pred <- 'dog_adolescent_or_adult_training'

for (outcome in outcomes) {
  cat(paste(replicate(80, '-'), collapse=''))
  cat(paste0('\n', outcome, '\n'))
  f <- as.formula(paste0(outcome, '~', '.'))
  
  df_tmp <- df_all[,c(outcome, pred)]
  #df_tmp <- apply_min_xtab(df_tmp, outcome)
  
  glm_fit <- glm(f, data=df_tmp, family='binomial')
  print(summary(glm_fit))
  print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
  #cat('\nVIF:\n')
  #print(car::vif(glm_fit))
  cat('\n')
}
```

Let's consider what other factors may also come in to play:

- Dog Age
- Dog Sex
- Acquiring the dog pre-adolescent
- Pre-adolescent training

We'll perform logistic regression to determine the impact of these factors. To
perform logistic regression we'll need to ensure out data subsets have enough
responses (`n >= 10`) for each possible answer to be included in the model.

```{r}
apply_min_xtab <- function(df, outcome, cutoff=10)
{
  drops <- NULL
  for (col in names(df)) {
    if (col == outcome) next
    if (is.integer(df[,col])) next
    if (is.numeric(df[,col])) next # added in order not to exclude dog age
    
    xtab <- table(df[,col], df[,outcome])
    if (min(xtab) < cutoff) {
      drops <- c(drops, col)
      break
    }
  }
  
  if (length(drops) > 0) {
    cat('\nDropped from model due to insufficient responses:\n')
    cat(drops)
    cat('\n')
  }
  
  return(df[, !(names(df) %in% drops)])
}
```

Now we perform the logistic regression.

when including age (the only continuous variable)
do not apply exclusion for min number of responses 

```{r}
pred <- 'dog_adolescent_or_adult_training'
glm_attribs <- c(
  'dog_age_in_months',
  'dog_gender',
  'dog_was_acquired_pre_adolescent',
  'dog_attended_pre_adolescent_training'
)
for (outcome in outcomes) {
  cat(paste(replicate(80, '-'), collapse=''))
  cat(paste0('\n', outcome, '\n'))
  f <- as.formula(paste0(outcome, '~', '.'))
  
  df_tmp <- df_valid_responses[,c(outcome, pred, glm_attribs)]
  #do not apply min function-- this only excludes age, which is continuous
  #df_tmp <- apply_min_xtab(df_tmp, outcome) 
  
  glm_fit <- glm(f, data=df_tmp, family='binomial')
  print(summary(glm_fit))
  print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
  cat('\nVIF:\n')
  print(car::vif(glm_fit))
  cat('\n')
}
```


# Investigating Training Specifics

we have identified that adolescent/adult training has a marginally 
significant impact on some behavior problems. 
Now we want to investigate the impact of certain
training factors.

## Isolating the Experimental Data Set

First, we isolate the experimental group (those who attended adolescent/adult
training) into their own data set.

```{r}
df_exp_all <- df_valid_responses %>%
  filter(dog_adolescent_or_adult_training == TRUE)
summary(df_exp_all)
```
#add some factors for training style
```{r}
df_exp_all <- df_exp_all %>%
  mutate(reward = ifelse(
    (is.na(dog_adolescent_or_adult_training_style) | 
      dog_adolescent_or_adult_training_style == 'I don\'t know'), NA, ifelse(
      dog_adolescent_or_adult_training_style == 'Reward based', TRUE, FALSE)))

df_exp_all <- df_exp_all %>%
  mutate(discipline = ifelse(
    (is.na(dog_adolescent_or_adult_training_style) |
      dog_adolescent_or_adult_training_style == 'I don\'t know'), NA, ifelse(
      dog_adolescent_or_adult_training_style == 'Discipline based', TRUE, FALSE)))

df_exp_all <- df_exp_all %>%
  mutate(mixed_style = ifelse(
    (is.na(dog_adolescent_or_adult_training_style) |
      dog_adolescent_or_adult_training_style == 'I don\'t know'), NA, ifelse(
      dog_adolescent_or_adult_training_style == 'Mixed balance of both reward and discipline', TRUE, FALSE)))
```

## Binary Logistic Regression

We want to answer the following questions about the training:

- Does continued training throughout dog's life produce a better outcome? 
- Did training technique (reward vs. punishment) have an impact on the outcome?
- Did the location of training have an impact on the outcome?

We will need to expand the independent variables used for the model to answer
these questions.

(The vast majority of dogs who had adolescent/adult training received 
reward-based training at group classes)

```{r}

outcomes_sig <- c(
  'dog_was_aggressive',
  'dog_has_fear_or_anxiety',
  'dog_has_jumped_problematically',
  'dog_has_barked_excessively',
  'dog_has_coprophagia',
  'dog_has_repetitive_behavior',
  'dog_has_house_soiled',
  'dog_has_rolled_in_repulsive_material',
  'dog_was_hyperactive',
  'dog_was_destructive',
  'dog_has_escaped',
  'dog_has_mounted'
)
outcomes_sig <- sort(outcomes_sig)

glm_attribs <- c(
  'dog_age_in_months',
  'dog_gender',
  'dog_was_acquired_pre_adolescent',
  'dog_attended_pre_adolescent_training'
)

#use single factor, reward or no
training_style <- c(
  'reward'
  #'discipline',
  #'mixed_style'
)
training_age <- c(
  'training_complete_age_6_to_18mo',
  'training_complete_age_18_to_36mo',
  'training_complete_age_continue_in_home',
  'training_complete_age_continue_classes'
)
training_location <- c(
  'trainer_group_classes',
  'trainer_boot_camp',
  'trainer_home_private'
)

glm_attribs <- c(
  glm_attribs,
  training_style,
  training_age,
  training_location
)

print(glm_attribs)

```

Now we can build and evaluate our models for the various outcomes.


```{r}

#note that training style appears to be highly co-variate with other variables
# when discipline and mixed style is included
# make it binary, reward or not reward


glm_attribs <- c(
  #glm_attribs,
  'dog_age_in_months',
  'dog_gender',
  'dog_was_acquired_pre_adolescent',
  'dog_attended_pre_adolescent_training',
  training_style,
  training_age,
  training_location
)

```


```{r}
for (outcome in outcomes_sig) {
  cat(paste(replicate(80, '-'), collapse=''))
  cat(paste0('\n', outcome, '\n'))
  f <- as.formula(paste0(outcome, '~', '.'))
  
  df_tmp <- df_exp_all[,c(outcome, glm_attribs)]
  df_tmp <- apply_min_xtab(df_tmp, outcome)
  
  # do not apply min xtab function to dog age
  
  # If necessary, drop columns due to separability problems.
  #if (outcome == 'destructive') {
  #  df_tmp <- subset(df_tmp, select=-c(shock_collar))
  #}
  
  glm_fit <- glm(f, data=df_tmp, family='binomial')
  print(summary(glm_fit))
  print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
  cat('\nVIF:\n')
  print(car::vif(glm_fit))
  cat('\n')
}
```
There seems to be very little impact of type of training, completion age, or training location
Pre-adolescent training was the most significant variable within the experimental group that predicts
presence of behavior

```{r}

```

