---
title: "Inferential Data Analysis"
author: "Liz Keller"
output: html_notebook
date: "`r Sys.Date()`"

---

```{r setup, include=FALSE}
library(tidyverse)
library(ggpubr)
library(GGally)
library(car)

knitr::opts_chunk$set(echo=TRUE)

#set.seed()
```

# Loading the Data

Load the raw data and verify its dimensions and structure.

```{r}
df <- readRDS('../data/tidy.Rds')
dim(df)
summary(df)

```

# subset only valid responses
# this is defined by dogs > 12 months of age, and respondent answered question whether dog had received adolescent/adult training

```{r}
df_valid_responses <- readRDS('../data/df_valid_responses.Rds')
dim(df_valid_responses)
summary(df_valid_responses)

#command to derive from full dataset:
#df_valid_responses <- subset(df, dog_adolescent_or_adult_training==TRUE | dog_adolescent_or_adult_training==FALSE)

```

# Control vs Experimental Groups

# with all valid responses:
control = adult training FALSE
exp = adult training TRUE


# using reward-only based adult training as exp
# (dogs who received discipline-based or mixed training excluded from exp group)
control = adult training FALSE
exp = reward_only_adult_training TRUE

```{r}

#drawing from all valid responses
df_control <- df_valid_responses %>%
  filter(dog_adolescent_or_adult_training == FALSE)
dim(df_control)

df_exp <- df_valid_responses %>%
  filter(dog_adolescent_or_adult_training == TRUE)
dim(df_exp)


# only dogs that received reward-based training
df_exp_reward <- readRDS('../data/df_reward_only_training.Rds')
dim(df_exp_reward)


```

## Overview

The first question we want to answer is if training during adolescence and/or 
adulthood has an impact on the likelihood of a dog having certain behavior problems. 

The behavior problems we are exploring are:

- Aggression
- Barking excessively
- Coprophagia
- Destructive behavior
- Escaping/running away
- Fear/anxiety
- House soiling
- Hyperactivity
- Jumping problematically
- Mounting
- Repetitive behavior
- Rolling in repulsive materials

```{r}
outcomes <- c(
  'dog_was_aggressive',
  'dog_has_fear_or_anxiety',
  'dog_has_jumped_problematically',
  'dog_has_barked_excessively',
  'dog_has_coprophagia',
  'dog_has_repetitive_behavior',
  'dog_has_house_soiled',
  'dog_has_rolled_in_repulsive_material',
  'dog_was_hyperactive',
  'dog_was_destructive',
  'dog_has_escaped',
  'dog_has_mounted'
)
outcomes <- sort(outcomes)
```

## Statistical Models

We check for correlation between the predictor (attending adolescent / adult training) and
each outcome (the presence of a specific behavior problem). 

Here we perform the Fisher Exact test with a Benjamini-Hochberg correction.
(The Fischer test is useful for categorical data where the sample size is small)

```{r}

pred <- 'dog_adolescent_or_adult_training'

idx <- NULL
p_values <- NULL
odds_ratios <- NULL 
conf_int_low <- NULL
conf_int_high <- NULL

for (outcome in outcomes) {
  tbl <- table(df_valid_responses[, pred], df_valid_responses[, outcome], dnn=c(pred, outcome))
  fisher <- fisher.test(tbl)
  idx <- c(idx, outcome)
  p_values <- c(p_values, fisher$p.value)
  odds_ratios <- c(odds_ratios, fisher$estimate[[1]])
  conf_int_low <- c(conf_int_low,fisher$conf.int[[1]])
  conf_int_high <- c(conf_int_high,fisher$conf.int[[2]])
}

# Correct for the possibility of Type I errors.
p_values <- p.adjust(p_values, method='BH')

# Form a result data frame.
df_out <- data.frame(outcome=idx, p_value=p_values,
                     odds_ratio=odds_ratios, 
                     conf_int1=conf_int_low, conf_int2=conf_int_high)

add_sig_columns <- function(df) {
  df$level <- ''
  df[df$p_value <= .05, 'level'] <- '*'
  df[df$p_value <= .01, 'level'] <- '**'
  df[df$p_value <= .001, 'level'] <- '***'
  
  df$dir <- ''
  df[df$odds_ratio < 1, 'dir'] <- '-'
  df[df$odds_ratio > 1, 'dir'] <- '+'
  
  return (df)
}

df_out <- add_sig_columns(df_out)
print(knitr::kable(df_out))

```
There is a significant negative correlation between adult 
training and some specific behaviors: house soiling and destructiveness

there is a significant positive correlation with adult training and
problematic jumping


#check chi-squared test for comparison

```{r}

pred <- 'dog_adolescent_or_adult_training'

idx <- NULL
p_values <- NULL
odds_ratios <- NULL 
for (outcome in outcomes) {
  tbl <- table(df_valid_responses[, pred], df_valid_responses[, outcome], dnn=c(pred, outcome))
  chisq <- chisq.test(tbl)
  idx <- c(idx, outcome)
  p_values <- c(p_values, chisq$p.value)
  #odds_ratios <- c(odds_ratios, chisq$estimate[[1]])
}

# Correct for the possibility of Type I errors (false positive/discovery rate).
p_values <- p.adjust(p_values, method='BH')

# Form a result data frame.
df_out_chisq <- data.frame(outcome=idx, p_value=p_values)

df_out_chisq <- add_sig_columns(df_out_chisq)
print(knitr::kable(df_out_chisq))

```
Result: same variables are of same order of significance with chi squared test


## Binary Logistic Regression

#The simplest possible model: adult / adolescent training as the sole predictor

```{r}
pred <- 'dog_adolescent_or_adult_training'

for (outcome in outcomes) {
  cat(paste(replicate(80, '-'), collapse=''))
  cat(paste0('\n', outcome, '\n'))
  f <- as.formula(paste0(outcome, '~', '.'))
  
  df_tmp <- df_valid_responses[,c(outcome, pred)]

  glm_fit <- glm(f, data=df_tmp, family='binomial')
  print(summary(glm_fit))
  print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
  #cat('\nVIF:\n')
  #print(car::vif(glm_fit))
  cat('\n')
}
```

Let's consider what other factors may also come in to play:

- Dog Age
- Dog Sex
- Acquiring the dog pre-adolescent
- Pre-adolescent (puppy) training
- interaction between puppy and adult training

We'll perform logistic regression to determine the impact of these factors. To
perform logistic regression we'll need to ensure out data subsets have enough
responses (`n >= 10`) for each possible answer to be included in the model.

```{r}
apply_min_xtab <- function(df, outcome, cutoff=10)
{
  drops <- NULL
  for (col in names(df)) {
    if (col == outcome) next
    if (is.integer(df[,col])) next
    if (is.numeric(df[,col])) next # added in order not to exclude dog age
    
    xtab <- table(df[,col], df[,outcome])
    if (min(xtab) < cutoff) {
      drops <- c(drops, col)
      break
    }
  }
  
  if (length(drops) > 0) {
    cat('\nDropped from model due to insufficient responses:\n')
    cat(drops)
    cat('\n')
  }
  
  return(df[, !(names(df) %in% drops)])
}
```

Now we perform the logistic regression.

when including age (the only continuous variable)
do not apply exclusion for min number of responses 

#without interaction terms

```{r}
pred <- 'dog_adolescent_or_adult_training'
glm_attribs <- c(
  'dog_age_in_months',
  'dog_gender',
  'dog_was_acquired_pre_adolescent',
  'dog_attended_pre_adolescent_training'
)

for (outcome in outcomes) {
  cat(paste(replicate(80, '-'), collapse=''))
  cat(paste0('\n', outcome, '\n'))
  f <- as.formula(paste0(outcome, '~', '.'))
  
  df_tmp <- df_valid_responses[,c(outcome, pred, glm_attribs)]
  #do not apply min function-- this only excludes age, which is continuous
  #df_tmp <- apply_min_xtab(df_tmp, outcome) 
  
  glm_fit <- glm(f, data=df_tmp, family='binomial')
  print(summary(glm_fit))
  print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
  cat('\nVIF:\n')
  print(car::vif(glm_fit))
  cat('\n')
}

# note VIF above 5 is red flag

```

# now try a model with an interaction term


```{r}


for (outcome in outcomes){
  
  #new section
  cat(paste(replicate(80, '-'), collapse=''))
  cat(paste0('\n', outcome, '\n'))
  
  #format variables as formula
  f <- as.formula(paste0(outcome, '~', 'dog_adolescent_or_adult_training  + dog_attended_pre_adolescent_training        + dog_adolescent_or_adult_training*dog_attended_pre_adolescent_training + dog_was_acquired_pre_adolescent        + dog_was_acquired_pre_adolescent*dog_attended_pre_adolescent_training + dog_age_in_months +  dog_gender'))
  

  df_tmp1 <- df_valid_responses[,c(outcome, 'dog_adolescent_or_adult_training', 'dog_attended_pre_adolescent_training','dog_was_acquired_pre_adolescent','dog_age_in_months','dog_gender')]

  df_tmp2 <- df_tmp1[complete.cases(df_tmp1),]

  glm_fit <- glm(f, data=df_tmp2, family = 'binomial')

  print(summary(glm_fit))
  
  print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
    cat('\nVIF:\n')
    print(car::vif(glm_fit,type="predictor"))
    cat('\n')

}



```



# instead of individual behaviors, use number of behavior problems as outcome

```{r}


#format variables as formula
f <- as.formula(paste0('num_behv_problems', '~', 'dog_adolescent_or_adult_training  + dog_attended_pre_adolescent_training  + dog_adolescent_or_adult_training*dog_attended_pre_adolescent_training + dog_was_acquired_pre_adolescent + dog_was_acquired_pre_adolescent*dog_attended_pre_adolescent_training + dog_age_in_months + dog_gender'))
  
  df_tmp1 <- df_valid_responses[,c('num_behv_problems', 'dog_adolescent_or_adult_training', 'dog_attended_pre_adolescent_training','dog_was_acquired_pre_adolescent','dog_age_in_months','dog_gender')]

df_tmp2 <- df_tmp1[complete.cases(df_tmp1),]

glm_fit <- glm(f, data=df_tmp2, family = 'quasi')

print(summary(glm_fit))

print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
  cat('\nVIF:\n')
  print(car::vif(glm_fit,type="predictor"))
  cat('\n')

```


# Investigating Training Specifics

we have identified that adolescent/adult training has a  
significant impact on jumping.

Does training style make a difference in the outcomes?
eliminate dogs who received discipline-based or mixed training from experimental group

```{r}

#identify dogs who received reward only adult training
df_reward_only_training <- subset(df_valid_responses, reward_only_adult_training==TRUE)

#combine control and modified exp groups
df_valid_responses_reward_only <- rbind(df_reward_only_training,df_control)

```


## repeat Binary Logistic Regression with reward-only experiment group


```{r}


for (outcome in outcomes){
  
  #new section
  cat(paste(replicate(80, '-'), collapse=''))
  cat(paste0('\n', outcome, '\n'))
  
  #format variables as formula
  f <- as.formula(paste0(outcome, '~', 'dog_adolescent_or_adult_training  + dog_attended_pre_adolescent_training  + dog_adolescent_or_adult_training*dog_attended_pre_adolescent_training + dog_was_acquired_pre_adolescent + dog_was_acquired_pre_adolescent*dog_attended_pre_adolescent_training + dog_age_in_months + dog_gender'))
  
  df_tmp1 <- df_valid_responses_reward_only[,c(outcome, 'dog_adolescent_or_adult_training', 'dog_attended_pre_adolescent_training','dog_was_acquired_pre_adolescent','dog_age_in_months','dog_gender')]

df_tmp2 <- df_tmp1[complete.cases(df_tmp1),]

glm_fit <- glm(f, data=df_tmp2, family = 'binomial')

print(summary(glm_fit))

print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
  cat('\nVIF:\n')
  print(car::vif(glm_fit,type="predictor"))
  cat('\n')

}



```
# instead of individual behaviors, use number of behavior problems as outcome

```{r}


#format variables as formula
f <- as.formula(paste0('num_behv_problems', '~', 'dog_adolescent_or_adult_training  + dog_attended_pre_adolescent_training  + dog_adolescent_or_adult_training*dog_attended_pre_adolescent_training + dog_was_acquired_pre_adolescent + dog_was_acquired_pre_adolescent*dog_attended_pre_adolescent_training + dog_age_in_months + dog_gender'))
  
  df_tmp1 <- df_valid_responses_reward_only[,c('num_behv_problems', 'dog_adolescent_or_adult_training', 'dog_attended_pre_adolescent_training','dog_was_acquired_pre_adolescent','dog_age_in_months','dog_gender')]

df_tmp2 <- df_tmp1[complete.cases(df_tmp1),]

glm_fit <- glm(f, data=df_tmp2, family = 'quasi')

print(summary(glm_fit))

print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
  cat('\nVIF:\n')
  print(car::vif(glm_fit,type="predictor"))
  cat('\n')

```



############################################################

Now we look just at the experiment group to see if style or location influenced the outcome

We want to answer the following questions about the training:

- Does continued training throughout dog's life produce a better outcome? 
- Did training technique (reward vs. punishment) have an impact on the outcome?
- Did the location of training have an impact on the outcome?

We will need to expand the independent variables used for the model to answer
these questions.

(The vast majority of dogs who had adolescent/adult training received 
reward-based training at group classes)

```{r}

outcomes_sig <- c(
  'dog_was_aggressive',
  'dog_has_fear_or_anxiety',
  'dog_has_jumped_problematically',
  'dog_has_barked_excessively',
  'dog_has_coprophagia',
  'dog_has_repetitive_behavior',
  'dog_has_house_soiled',
  'dog_has_rolled_in_repulsive_material',
  'dog_was_hyperactive',
  'dog_was_destructive',
  'dog_has_escaped',
  'dog_has_mounted'
)
outcomes_sig <- sort(outcomes_sig)

#use single factor, reward or no
training_style <- c(
  'reward_only_puppy_training',
  'reward_only_adult_training'
  #'discipline',
  #'mixed_style'
)
training_age <- c(
  #'training_complete_age_6_to_18mo',
  #'training_complete_age_18_to_36mo',
  'training_complete_age_continue_in_home',
  'training_complete_age_continue_classes'
)
training_location <- c(
  'puppy_group_classes', 
  'trainer_group_classes'
)

glm_attribs <- c(
  training_style,
  training_age,
  training_location
)

print(glm_attribs)

```

# Build and evaluate models for the various outcomes.

```{r}

# we don't need to include adult training as variable because by definition all dogs in experimental group have had adult training
# There will be some cases where there are less than 10 responses for a given variable/outcome combination, so eliminate those variables from the model using apply_min_xtab function defined above
# Puppy training style and location creates problems in some models due to small sample size / multi colinearity

for (outcome in outcomes){
  
  #new section
  cat(paste(replicate(80, '-'), collapse=''))
  cat(paste0('\n', outcome, '\n'))
  
  #format variables as formula
  #f <- as.formula(paste0(outcome, '~', 'reward_only_puppy_training + reward_only_adult_training + puppy_group_classes + trainer_group_classes + training_complete_age_continue_in_home + training_complete_age_continue_classes'))
  f <- as.formula(paste0(outcome, '~', '.'))
  
  #df_tmp1 <- df_exp[,c(outcome, 'reward_only_puppy_training', 'reward_only_adult_training','puppy_group_classes', 'trainer_group_classes', 'training_complete_age_continue_in_home', 'training_complete_age_continue_classes')]
  df_tmp1 <- df_exp[,c(outcome, 'reward_only_adult_training','trainer_group_classes', 'training_complete_age_continue_in_home', 'training_complete_age_continue_classes')]

df_tmp2 <- df_tmp1[complete.cases(df_tmp1),]
df_tmp2 <- apply_min_xtab(df_tmp2, outcome) 

glm_fit <- glm(f, data=df_tmp2, family = 'binomial')

print(summary(glm_fit))

print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
  cat('\nVIF:\n')
  print(car::vif(glm_fit))#,type="predictor"))
  cat('\n')

}



```

# instead of individual behaviors, use number of behavior problems as outcome

```{r}


#format variables as formula

f <- as.formula(paste0('num_behv_problems', '~', '.'))
  
#df_tmp1 <- df_exp[,c(outcome, 'reward_only_puppy_training', 'reward_only_adult_training','puppy_group_classes', 'trainer_group_classes', 'training_complete_age_continue_in_home', 'training_complete_age_continue_classes')]
df_tmp1 <- df_exp[,c('num_behv_problems', 'reward_only_adult_training','trainer_group_classes', 'training_complete_age_continue_in_home', 'training_complete_age_continue_classes')]
  
df_tmp2 <- df_tmp1[complete.cases(df_tmp1),]

glm_fit <- glm(f, data=df_tmp2, family = 'quasi')

print(summary(glm_fit))

print(exp(cbind(OR=coef(glm_fit), suppressMessages(confint(glm_fit)))))
  cat('\nVIF:\n')
  print(car::vif(glm_fit,type="predictor"))
  cat('\n')

```


There seems to be very little impact of style of training, completion age, or training location on behavior outcomes



```{r}


```

