---
title: "Exploratory Data Analysis"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggpubr)
library(GGally)
library(moments)
library(scales)
knitr::opts_chunk$set(echo=TRUE)
```

# Overview

This notebook documents the exploratory data analysis undertaken for this
project.

# Loading the Data

Load the tidy data set into memory. If it does not exist, then run the notebook
that builds it.

```{r}
tidy_data_filepath <- '../data/tidy.Rds'
if (!file.exists(tidy_data_filepath)) {
  rmarkdown::render('./notebook-0-tidy.Rmd')
}
df <- readRDS(tidy_data_filepath)
stopifnot(identical(dim(df)+0, c(1011, 111)))
str(df)
```
## Overview of Data Set

# summarize data

```{r}
summary(df)

```

```{r}
# Number of unique owners after inclusion criteria.
length(unique(df$owner_id))

```

```{r}
#number of dogs per household.
summary(plyr::count(df, 'owner_id'))
```
Median number of dogs per household = 1 (range: 1 to 21), mean 1.6. Now we can
drop the column to simply the data set.

```{r}
df <- subset(df, select=-c(owner_id))
```

Before we look at the data, let's add a basic behavior problem indicator column.

#how may dogs have any behavior issue? 

```{r}

df <- df %>%
  mutate(behav_problem = ifelse(
    dog_was_aggressive | dog_has_fear_or_anxiety | dog_has_jumped_problematically | 
      dog_has_barked_excessively | dog_has_coprophagia | dog_has_repetitive_behavior |
      dog_has_house_soiled | dog_has_rolled_in_repulsive_material | dog_was_hyperactive | 
      dog_was_destructive | dog_has_escaped | dog_has_mounted, TRUE, FALSE))
summary(df$behav_problem)

#create variable with number of behavior problems
df_temp <- df[,c('dog_was_aggressive', 'dog_has_fear_or_anxiety', 'dog_has_jumped_problematically', 
      'dog_has_barked_excessively', 'dog_has_coprophagia', 'dog_has_repetitive_behavior',
      'dog_has_house_soiled', 'dog_has_rolled_in_repulsive_material', 'dog_was_hyperactive', 
      'dog_was_destructive', 'dog_has_escaped', 'dog_has_mounted')]

#count number of 'True' responses per row
df$num_behv_problems <- rowSums(df_temp,na.rm=TRUE)
summary(df$num_behv_problems)
  
```

# Summary stats specific to only valid responses which meet the inclusion criteria (valid response = dog > 12 months of age, answered True or False to dog_adolescent_or_adult_training)

```{r}

df_valid_responses <- subset(df, dog_adolescent_or_adult_training==TRUE | dog_adolescent_or_adult_training==FALSE)
summary(df_valid_responses)

```

```{r}

#summary of number of behavior problems 
summary(df_valid_responses$num_behv_problems)

#number of dogs that engage in continued training
length(which(df_valid_responses$training_complete_age_continue_in_home==TRUE | df_valid_responses$training_complete_age_continue_classes==TRUE))

```


#look at distribution of number of behavior problems

```{r}

p <- ggplot(df_valid_responses, aes(x=num_behv_problems)) + geom_bar(fill='steelblue') + 
  scale_x_continuous(breaks=0:10) + theme(axis.text=element_text(size=14),    axis.title=element_text(size=18,face="bold")) # + geom_text(aes(label = after_stat(count)), stat = "count", vjust = 1.1, colour = "white")  

p.labs <- p + labs(x = "Number of behavior problems per dog", y="Count") 
p.labs

```


```{r} 
#save figure
filename = "/Users/ekeller/ccbs/ccbs-study-3/figures/fig1_number_of_behavior_problems.pdf"

pdf(file=filename, width=12, height=8)
p.labs
dev.off()
```


#calculate percentage of dogs with each behavior problem

```{r}

behav_probs <- c(
  'dog_was_aggressive',
  'dog_has_fear_or_anxiety',
  'dog_has_jumped_problematically',
  'dog_has_barked_excessively',
  'dog_has_coprophagia',
  'dog_has_repetitive_behavior',
  'dog_has_house_soiled',
  'dog_has_rolled_in_repulsive_material',
  'dog_was_hyperactive',
  'dog_was_destructive',
  'dog_has_escaped',
  'dog_has_mounted'
)
behav_probs <- sort(behav_probs)

tmp <- df_valid_responses %>% select(all_of(behav_probs))

```

```{r}

#create table for percentage occurence

# Define a dataframe 
pct_df <- data.frame(Behavior  = behav_probs,
                  Occurence = numeric(length(behav_probs)))

for (behav_prob in behav_probs){
  
  tbl <- plyr::count(tmp, behav_prob) / nrow(tmp) * 100
  #tmp <- tmp %>% 
  ## add percentage label with `sprintf()`
  #perc = paste0(sprintf("%4.1f", tbl$freq[2]), "%")
  perc = tbl$freq[2]
  
  pct_df[pct_df$Behavior==behav_prob,2] = perc
  
}

pct_df <- pct_df[order(pct_df$Occurence,decreasing = TRUE),] 


```

```{r}

my_labels <- c("Excessive Barking","Coprophagia","Escaping", "Fear or Anxiety", "House soiling",               "Problematic jumping", "Mounting", "Repetitive Behavior", "Rolling in Repulsive Material", "Aggression","Destructive", "Hyperactivity" )

pct_labels = paste0(sprintf("%4.1f", pct_df$Occurence), "%")

df_sorted <- arrange(pct_df, Occurence) 

p <- ggplot(pct_df, aes(x=Behavior,y=Occurence)) + geom_bar(stat="identity",fill='steelblue') + scale_x_discrete(labels=my_labels) + scale_y_continuous(limits=c(0,100)) + geom_text(aes(label = pct_labels), hjust = -0.2, colour = "black", size=5)  + theme(axis.text=element_text(size=14),    axis.title=element_text(size=16,face="bold")) 

p.labs <- p + coord_flip() + labs(y = "percentage of dogs", ) 
p.labs


```
```{r} 
#save figure
filename="/Users/ekeller/ccbs/ccbs-study-3/figures/behavior_problems_pct.pdf"
pdf(file=filename, width=12, height=8)
p.labs
dev.off()

```


# general notes on dataset: (all stats exclude no answer or NA's unless otherwise stated)
-- 322 (out of 1011 total obs) responses are excluded due to no response (NA's) in either training or behavior questions; many of these dogs were too young to take part in the survey (<12 mo)
-- 671 out of 689 or 97.4% reported at least one behavior problem
-- median number of behavior problems is 4, and mean is 3.7
-- Dog gender spilt almost even, slight bias towards male (478 F / 507 M total, 325 F / 364 M with valid responses)
-- 634 unique dog owners, median 1 dog owned, mean 1.6 (all respondents)
-- Mean age (in months) is 80.9 (6.7 years) and median age is 75 (6.25 years) [Note all dogs less than 12 months excluded from analysis]
-- a majority of respondents (389/689 = 56%) acquired their dog in the pre-adolescent stage
-- of the respondents who provided an answer (559), just under half of dogs had attended pre-adolescent training (260/559 = 46.5%). The majority of these dogs (209/260 = 80.4%) had received reward-based training only. A further 15% (39/260) received a mix of reward and discipline-based training. The vast majority (229/260 = 88%) attended group classes (as opposed to private training or boot camp).
-- More than half of dogs had attended adolescent or adult training (442/689 = 64.2%). The vast majority (358/442 = 81.0%) reported receiving reward-based training only and attended group classes (387/442 = 87.6%)

-- the majority continued some form of training post-puppyhood: 11% (77/689) completed in adolescence (defined as 6 to 18 months), 10.6% (73/689) completed between ages 18-36 months (social maturity), 39.6% (273/689) continue to train in-home, and 32.8% (226/689) continue to attend classes, with 53% (365/689) reporting continued training either in home or through classes 


#specific behavior problems:
-- 56.2% reported aggression (387/689)
-- 73.9% reported fear or anxiety (509/689)
-- 19.4% reported problematic jumping (134/689)
-- 19.0% reported excessive barking (131/689)
-- 41.7% reported coprophagia (287/689)
-- 26.9% reported repetitive behavior (185/689)
-- 23.1% reported house soiling (159/689)
-- 41.1% reported rolling in repulsive material (283/689)
-- 12.5% reported hyperactivity (86/689)
-- 15.4% reported destructive behavior (106/689)
-- 26.3% reported escaping (181/689)
-- 18.4% reported mounting (127/689)


# make some plots to visualise data

How many of dogs with behavior problems have had adolescent/adult training?

```{r}

ggplot(df_valid_responses, aes(x = dog_adolescent_or_adult_training)) +
  geom_bar() + geom_text(aes(label=..count..),stat='count',vjust=-0.2)

ggplot(data = df_valid_responses) +  
       geom_count(mapping = aes(x = dog_adolescent_or_adult_training, y = behav_problem)) 

df_valid_responses %>% 
  # create a column called n for number of times x & y occur together
  count(dog_adolescent_or_adult_training, behav_problem) %>% 
  # create columns for each unique value of y and 
  # put the values of column n below
  spread(key = behav_problem, value = n, fill = 0) 

df_valid_responses %>% 
  count(dog_adolescent_or_adult_training, behav_problem) %>%  
  ggplot(mapping = aes(x = dog_adolescent_or_adult_training, y = behav_problem)) +
    geom_tile(mapping = aes(fill = n))

```

Almost all dogs have at least one behavior problem, so this is not very meaningful


# examine variable distribution for all variables

```{r}

# all responses
ggplot(df, aes(dog_age_in_months)) + geom_bar(fill='steelblue')
skewness(df$dog_age_in_months,na.rm=TRUE)

ggplot(df, aes(owner_age)) + geom_bar(fill='steelblue')
skewness(df$owner_age,na.rm=TRUE)

#valid responses only
ggplot(df_valid_responses, aes(dog_age_in_months)) + geom_bar(fill='steelblue')
skewness(df_valid_responses$dog_age_in_months,na.rm=TRUE)

ggplot(df_valid_responses, aes(owner_age)) + geom_bar(fill='steelblue')
skewness(df_valid_responses$owner_age,na.rm=TRUE)

```

We see a modest right skew (younger) in the dog age plot, and a left skew (older) in the owner age plot. Let's try to center it by applying a log transform.

```{r}
df_valid_responses <- df_valid_responses %>%
  mutate(log_dog_age = log(dog_age_in_months))
ggplot(df_valid_responses, aes(log_dog_age)) + geom_bar(fill='steelblue')
skewness(df_valid_responses$log_dog_age,na.rm=TRUE)


df_valid_responses <- df_valid_responses %>%
  mutate(log_owner_age = log(owner_age))
ggplot(df_valid_responses, aes(log_owner_age)) + geom_bar(fill='steelblue')
skewness(df_valid_responses$log_owner_age,na.rm=TRUE)

df_valid_responses <- subset(df_valid_responses, select=-c(log_dog_age))
df_valid_responses <- subset(df_valid_responses, select=-c(log_owner_age))

```
Dog and owner age still skews with log transform -- discard this column


# We want to isolate the effect of adolescent / adult training

 -- using all valid answers which meet our inclusion criteria = df_valid_responses
 -- controlling for the fact that some dogs have also received pre-adolescent training:
    control group: those who had pre-adolescent training but no further training
    exp group: those with pre-adolescent training and adolescent and/or adult training

```{r} 
#testing some different groupings of the data

#separate by puppy training
df_control_puppy_training_only <- subset(df, dog_attended_pre_adolescent_training==TRUE & dog_adolescent_or_adult_training==FALSE)
df_exp_puppy_and_adult_training <- subset(df, dog_attended_pre_adolescent_training==TRUE & dog_adolescent_or_adult_training==TRUE)
df_adult_only <- subset(df, dog_attended_pre_adolescent_training==FALSE & dog_adolescent_or_adult_training==TRUE)
df_no_training <- subset(df, dog_attended_pre_adolescent_training==FALSE & dog_adolescent_or_adult_training==FALSE)

nrow(df_control_puppy_training_only) + nrow(df_exp_puppy_and_adult_training) + nrow(df_adult_only) + nrow(df_no_training)

#subset of data with dogs that continue training
df_continue_training <- subset(df, training_complete_age_continue_in_home==TRUE | training_complete_age_continue_classes==TRUE)

#subset of data to eliminate dogs that received some discipline-based training
df_reward_only_training <- subset(df_valid_responses, reward_only_adult_training==TRUE)
df_control_no_adult_training <- subset(df_valid_responses, dog_adolescent_or_adult_training==FALSE)

nrow(subset(df_reward_only_training))
nrow(subset(df_control_no_adult_training))

```

# Including the effect of pre-adolescent training:
Of the 1011 total survey responses, 452 are excluded because they did not answer whether the dog had received pre-adolescent training and/or adult training. This leaves 559 responses. The breakdown of the group:

177 had NO training (neither pre-adolescent or adult)
37 had pre-adolescent but NO further training
223 had pre-adolescent AND adult training
122 had NO pre-adolescent training but did have adult training


# Controlling for the adverse impact of discipline-based training

Looking at reward-only based training vs. no training (eliminating dogs that had received discipline-based or mixed discipline and reward training):
605 responses in dataset
358 received reward-only adult training (experiment)
247 did not receive adult training (control)


# How many dogs with any type of training (either puppy or adult) have behavior problems? 

```{r}

# Create column indicating whether the dog had any type of training (pre-adolescent or adolescent/adult)

df_valid_responses <- df_valid_responses %>%
  mutate(any_training = ifelse(
    dog_attended_pre_adolescent_training | dog_adolescent_or_adult_training & !is.na(dog_attended_pre_adolescent_training), 
    TRUE, FALSE))
summary(df_valid_responses$any_training)

```


```{r}

df1 = subset(df_valid_responses, any_training == FALSE & behav_problem == FALSE)
df2 = subset(df_valid_responses, any_training == FALSE & behav_problem == TRUE)

df3 = subset(df_valid_responses, any_training == TRUE & behav_problem == FALSE)
df4 = subset(df_valid_responses, any_training == TRUE & behav_problem == TRUE)

```

6 dogs with no training have no behavior problems
171 dogs with no training do have behavior problems
11 dogs with some training have no behavior problems
371 dogs with some training do have behavior problems

Again, most dogs in the study have at least one behavior problem so this is not very insightful



the majority of dogs with behavior problems have between 2-5 separate behavior issues

Is there a correlation between dog age and behavior problems?

```{r}

data<-dplyr::select(df_valid_responses,num_behv_problems,dog_age_in_months)
ggpairs(data)

```

There does not appear to be a significant correlation between a dog's age and behavior problems, except perhaps at the extreme end (the dogs with 9 or 10 problems tend to be younger)

#visualise continuous variables (dog age and num of behavior problems)

```{r}
df_valid_responses %>%
  ggpairs(columns=c('dog_age_in_months', 
                    'num_behv_problems'),
          mapping=ggplot2::aes(color=dog_attended_pre_adolescent_training),
          diag=list(discrete='barDiag',
                    continuous=wrap('densityDiag', alpha=0.5)),
          legend=1,
          progress=FALSE) +
  theme(legend.position='bottom')


df_valid_responses %>%
  ggpairs(columns=c('dog_age_in_months', 
                    'num_behv_problems'),
          mapping=ggplot2::aes(color=dog_adolescent_or_adult_training),
          diag=list(discrete='barDiag',
                    continuous=wrap('densityDiag', alpha=0.5)),
          legend=1,
          progress=FALSE) +
  theme(legend.position='bottom')
```
Older dogs are more likely to have received some adult training. 

Dogs that have received adult training seem to have fewer problems on average

```{r}

#calculate mean #of problems for each group
df1 <- subset(df_valid_responses, dog_adolescent_or_adult_training==FALSE)
df2 <- subset(df_valid_responses, dog_adolescent_or_adult_training==TRUE)

mean(df1$num_behv_problems)
mean(df2$num_behv_problems)

median(df1$num_behv_problems)
median(df2$num_behv_problems)

```

## Discrete Variables

### Independent Variables

```{r, fig.height=5, fig.width=5}
vars <- c(
  'any_training',
  'dog_was_acquired_pre_adolescent',
  'dog_attended_pre_adolescent_training',
  'dog_adolescent_or_adult_training',
  'dog_gender',
  'trainer_group_classes',
  'trainer_boot_camp',
  'trainer_home_private',
  'training_complete_age_6_to_18mo',
  'training_complete_age_18_to_36mo',
  'training_complete_age_continue_in_home',
  'training_complete_age_continue_classes'
)

plot_list <- list()
for (i in 1:length(vars)) {
  col <- vars[i]
  p <- df_valid_responses %>%
    select(col) %>%
    drop_na(col) %>%
    ggplot(aes_string(x = col)) +
    geom_bar(fill='steelblue')
  plot_list[[i]] <- p
}
ggarrange(plotlist=plot_list, ncol=3, nrow=4)

```
```{r} 
#save figure
png(file="/Users/ekeller/ccbs/ccbs-study-3/figures/training_variables_barplot.png",
width=12, height=8, units="in", res=300)
ggarrange(plotlist=plot_list, ncol=3, nrow=4)
dev.off()
```

### Dependent Variables

```{r, fig.height=4, fig.width=5}
outcomes <- c(
  'dog_was_aggressive',
  'dog_has_fear_or_anxiety',
  'dog_has_jumped_problematically',
  'dog_has_barked_excessively',
  'dog_has_coprophagia',
  'dog_has_repetitive_behavior',
  'dog_has_house_soiled',
  'dog_has_rolled_in_repulsive_material',
  'dog_was_hyperactive',
  'dog_was_destructive',
  'dog_has_escaped',
  'dog_has_mounted'
)

plot_list <- list()
for (i in 1:length(outcomes)) {
  col <- outcomes[i]
  p <- df_valid_responses %>%
    select(col) %>%
    drop_na(col) %>%
    ggplot(aes_string(x = col)) +
    geom_bar(fill='steelblue')
  plot_list[[i]] <- p
}
ggarrange(plotlist=plot_list, ncol=4, nrow=3)
```

```{r} 
#save figure
png(file="/Users/ekeller/ccbs/ccbs-study-3/figures/behavior_problems_barplot.png",
width=12, height=8, units="in", res=300)
ggarrange(plotlist=plot_list, ncol=3, nrow=4)
dev.off()
```





```{r}

#save all valid responses as dataset
saveRDS(df_valid_responses, '/Users/ekeller/ccbs/ccbs-study-3/data/df_valid_responses.Rds')


#save datasets separated by training style 
saveRDS(df_reward_only_training, '/Users/ekeller/ccbs/ccbs-study-3/data/df_reward_only_training.Rds')
saveRDS(df_control_no_adult_training,'/Users/ekeller/ccbs/ccbs-study-3/data/df_control_no_adult_training.Rds')


```